Typed Interactions for NLU in Opal

### 2017-12-11 paper.tex
Added content, yay

\usepackage{minted}
\usepackage{graphicx}
\usepackage{titlesec}
\usepackage[english]{babel}
\usepackage{enumitem}
\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry}
\usepackage[bottom]{footmisc}
\usepackage[hidelinks]{hyperref}
\usepackage{amsmath}
\usepackage{fixme}
\usepackage{xcolor}

@@

\section{Formalism} \label{formalism}
In order to reason about the correctness of our translations, we have formalized
the results in section \ref{description}. We begin by formalizing the type DSL
as a pseuo-BNF grammar. This is shown in figure \ref{fig:grammar}.

\begin{figure}
  \centering
  \begin{align*}
    \fcy{D} ::=&\ \text{FreeText}(\etag{t}) \\
    |&\ \text{Keywords}(\etag{t},\ [\ell]) \\
    |&\ \text{Trait}(\etag{t},\ \{s: \tau\}) \\
    |&\ \text{Alias}(\etag{t},\ \tau) \\
    \\
    \tau ::=&\ \etag{t} \mid \ell \mid \{s: \tau\} \mid \dots
  \end{align*}
  $$ \etag{t} \in \textsf{Tag} \qquad \ell \in \textsf{Literal} \qquad s \in
  \textbf{string} $$
  \caption{A grammar for the DSL.}
  \label{fig:grammar}
\end{figure}

The translation from the haskell declarations above is mostly straightforward,
but there are a few notable differences. First, we have separated out tags,
literals, and strings. This is mostly for notational convenience, but these
things are also slightly different when serialized to Typescript. Tags become
names in Typescript, identifying a specific type alias. Literals become strings
in a literal type. Finally, strings become simple strings, used within a record
as keys.

We will define a set $\textsf{Entity} \subseteq \textsf{Tag}$ to represent the
available entity tags. Any of these might be needed to reconstruct the
appropriate type from the response.

Given a tag \etag{t}, we can define a parser in terms of \etag{t} as follows:
\begin{align*}
  \text{interp}\ &:\ \textsf{Tag} \rightarrow \fcy{T} \\
  \text{parse}\ &:\ 2^{\textsf{Entity}} \rightharpoonup \text{interp}(\etag{t})
\end{align*}
The interp map computes the Typescript type of \etag{t}. Then, we can represent
parsing as a partial function from sets of entities to typescript objects of the
correct type.

\section{Implementation} \label{implementation}
Explain the architecture of the implementation. Discuss how the tools should be
used.

>>>>

\usepackage{minted}
\usepackage{graphicx}
\usepackage{titlesec}
\usepackage[english]{babel}
\usepackage{enumitem}
\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry}
\usepackage[bottom]{footmisc}
\usepackage[hidelinks]{hyperref}
\usepackage{amsmath}
\usepackage{fixme}
\usepackage{xcolor}
\usepackage{subcaption}

@@

\section{Formalism} \label{formalism}
In order to reason about the correctness of our translations, we have formalized
the results from section \ref{description}. We begin by formalizing the type DSL
as a pseuo-BNF grammar. This is shown in figure \ref{fig:grammar_a}.
\begin{figure}
  \centering
  \begin{subfigure}{1\linewidth}
    \begin{align*}
      \fcy{D} ::=&\ \text{FreeText}(\etag{t}) \\
      |&\ \text{Keywords}(\etag{t},\ [\ell]) \\
      |&\ \text{Trait}(\etag{t},\ \{s: \fcy{T}\}) \\
      \\
      \fcy{T} ::=&\ \fcy{D} \mid \ell \mid \{s: \fcy{T}\}
    \end{align*}
    $$ \etag{t} \in \textsf{Tag} \qquad \ell, s \in \textbf{string} $$
    \caption{Type DSL}
    \label{fig:grammar_a}
  \end{subfigure}\vspace{1cm}
  \begin{subfigure}{1\linewidth}
    $$ \tau ::=\ \textbf{string} \mid \ell \mid \{s: \tau\} \mid \langle \ell
    \rangle $$
    \caption{Relevant Opal Types}
    \label{fig:grammar_b}
  \end{subfigure}
  \caption{Formal grammars.}
\end{figure}
There are similarities between these definitions and the haskell ones in figure
\ref{fig:ast}, but it has some slightly different goals. First, we remove
aliases and require that all type definitions are written ``in-line''. Next,
rather than split the definition by allowing tags within $\fcy{T}$, we also
require that entitiy definitions are inlined. This means that the data a user
expects from the NLU backend can be represented by a single tree of type
$\fcy{D}$. Importantly, none of these simplifications are fundamental---they
simply remove some aspects of the system that make it easier to use in practice.

In addition to a grammar for the DSL itself, in figure \ref{fig:grammar_b} we
define a grammar of types, $\tau$ which will represent Opal types specifically.
We can think of $\tau$ as representing the final form of the data that a user
hopes to obtain, with no information about the entities that generated the data.
We add a few types that are noteworthy: the type {\bf string} is added to
account for FreeText entities; $\langle {\ell} \rangle$ represents a disjunction
of literals, and covers the case of Keywords entities.

As mentioned above, user's view of their system can be formalized as a top-level
declaration generated by $\fcy{D}$. Since this declaration encompases both type
and configuration information, we would like a way to compute just the
\emph{type} of data that the user expects from the NLU engine. We define the
following mutually recursive functions, which traverse a declaration and
construct a single type in $\tau$.
\begin{align*}
  \text{interp}\ &:\ \fcy{D} \to \tau \\
  \text{interpTy}\ &:\ \fcy{T} \to \tau
\end{align*}

With the basic building blocks defined, we can move forward in formalizing our
system. \fxnote{Should I write out implementations?}

\subsection{Configuration Generation}
The first step that our system takes is configuration. In practice, this is a
two-part process: the user needs to extract the appropriate Opal code as well as
the configuration for the NLU backend. The first step is covered by our
definition of interp. Since we are taking $\tau$ as an analog for relevant Opal
types, the type generation process is exactly the execution of interp on the
top-level entity declaration.

As for the second step, we would like the NLU backend to be set up such that its
responses are \emph{well-formed} with respect to the configuration. Intuitively,
any time we look up some entity tag in the response, we would like to get data
that is compatible with that entity's declaration. We define a type,
\textsf{Resp} of responses, and require that a lookup function exists, with the
following type:
$$ \text{lookup}\ :\ \textsf{Tag} \to \textsf{Resp} \rightharpoonup
\textbf{string} $$
Note that the function is partial; there is no guarantee that the response will
contain the desired data. In the sequel, we will often assume that lookup is
defined on tags we care about, but that assumption will be stated explicitly
rather than assumed by default.

The lookup function always returns a string, but the string might mean any
number of things. The data might be a particular literal, in the case of a
keywords or trait entity, or an entire phrase in the case of a free-text entity.
We define configurations as functions that map a subset of entity tags to the
type of data expected for those entities.
$$ \textsf{Cfg} \equiv \textsf{Tag} \rightharpoonup \tau $$
Now, we call a response well-formed with respect to $\text{cfg} : \textsf{Cfg}$
if and only if it returns appropriate data on lookup. More formally, for some $r
: \textsf{Resp}$,
\begin{align*}
  &r\ \text{well-formed}_{\text{cfg}} \iff \\
  &\forall (\etag{t} : \textsf{Tag}).\ \text{cfg}(\etag{t})\ \text{defined} \Rightarrow \text{lookup}(\etag{t}, r) : \text{cfg(\etag{t})}
\end{align*}

\subsection{Response Parsing}
We discussed above that there is a challenge in parsing the unstructured
response from the NLU engine into data of the appropriate types. This section
will formalize that problem and present our solution.

Our goal will be to write a parsing function which takes a declaration, $d$, and
a response, and produces a value of type interp($d$). We will again write two
mutually recursive functions, parse and parseTy, with types:
\begin{align*}
\text{parse}\ &:\ \Pi_{(d : \fcy{D})}. \textsf{Resp} \rightharpoonup \text{interp}(d) \\
\text{parseTy}\ &:\ \Pi_{(t : \fcy{T})}. \textsf{Resp} \rightharpoonup \text{interp}(t)
\end{align*}
Since we need the type of the result to depend on the declaration, these are
dependent functions. The functions are also partial because lookups might fail.

We will define parse by case analysis on the declaration.
\begin{align*}
  \text{parse}(\text{FreeText}(\etag{t}), r) = \text{lookup}(\etag{t}, r)
\end{align*}
This case is fairly straightforward. Since the interpretation of a free-text
entity is just a string, we can simply look up the value of the appropriate
entity from the response.
\begin{align*}
  \text{parse}(\text{Keywords}(\etag{t}, [\ell_1, \dots, \ell_n]), r) =\\
  \text{lookup}(\etag{t}, r)
\end{align*}
Perhaps surprisingly, we can use the exact same definition in the Keywords case
that we did in the FreeText case. We are required to return a value of type
$\langle {\ell_1 \mid \dots \mid \ell_n} \rangle$, but since the NLU engine has
been configured with the same list of literals, we are guaranteed to get one of
them back from lookup (assuming lookup is defined).
\begin{align*}
  \text{parse}(\text{Trait}(\etag{t}, rec), r) = \\
  \text{parseTy}(rec[\text{lookup}(\etag{t}, r)], r)
\end{align*}
The final case for parse simply matches the value in the response to an index
into the record, and interprets that record entry using parseTy.

We will not write out the full definition of parseTy, but note that it is fairly
automatic. If the type is another declaration, we can call back to parse, if it
is a literal we simply generate the appropriate string, and if it is a record,
we build the appropriate record and recursively build out its fields.

\subsection{Correctness}
Given a declaration $d$, we generate a configuration, cfg. Then,
\begin{flalign*}
  &\forall\ (r : \textsf{Resp}).\\
  &r\ \text{well-formed}_{\text{cfg}} \Rightarrow \text{parse}(d, r)\ \text{defined}
\end{flalign*}

\section{Implementation} \label{implementation}

\subsection{Configuration Tool}
When a user wants to configure an Opal application to interact with an NLU
library like Wit, they begin by using our configuration tool. The tool turns
code in the type DSL into typescript code and JSON configuration, and ensures
that the NLU engine responses will agree with the type definitions in the user's
application.

Our configuration tool is written in Haskell, and is designed to be simple and
obviously correct. We use the Parsec library to parse the type DSL into the
intermediate representation mentioned above in figure \ref{fig:ast}.

\subsection{Runtime Library}
